### 2023-07-16 tex/syntax.tex
[ writing ] address all remaining todos

\section{Terms of a \AgdaRecord{System}}\label{sec:terms}

@@

To specify the meaning of descriptions, we assume some \AgdaBound{X} : \AgdaFunction{ExtOpenFam},
% \ExecuteMetaData[\Interpretationtex]{X},
over which we form one layer of syntax, using the function
\AgdaFunction{$\llbracket$\_$\rrbracket$p} that interprets
\AgdaDatatype{Premises} defined below.  The first argument to
\AgdaBound{X} is the new variables bound by this layer of syntax, as
exemplified in the first clause of
\AgdaFunction{$\llbracket$\_$\rrbracket$p}.  The second argument is
the context containing the variables being carried over from the
previous layer.  Notice that this is not, in general, the same as the
context from the previous layer, because the usage annotations may
have been changed by connectives like
\AgdaInductiveConstructor{\_`$*$\_} and
\AgdaInductiveConstructor{\_`$\cdot$\_}.  The third argument is the
type of subterm required.

The remainder of the clauses of \AgdaFunction{$\llbracket$\_$\rrbracket$p}
are the interpretation into bunched combinators.
The superscript \AgdaFunction{$^c$} on the bunched connectives denotes that
they have been lifted from predicates on usage vectors to predicates on
contexts, with the type component of the context shared throughout.
Additive connectives \AgdaFunction{$\dot1$} and \AgdaFunction{$\dottimes$} are
already polymorphic (not relying on anything specific about usage vectors), so
do not need a \AgdaFunction{$^c$} variant.

@@

The most obvious way to make such an \AgdaBound{X} is to use some existing
\AgdaFunction{OpenFam} on an extended context.
We defined \AgdaFunction{Scope} to do this: take the new variables
\AgdaBound{$\Delta$}, concatenate them onto the existing context
\AgdaBound{$\Gamma$}, and pass the extended context onto the judgement
\AgdaBound{T}.

@@

We use \AgdaFunction{Scope} to deal with new variables in syntax.
Terms resemble the free monad over a layer-of-syntax functor, though
that picture is complicated by variable binding.  A term is either a
variable or a use of a logical rule together with terms for each of
the required subterms. The \AgdaFunction{Size} argument is a use of
Agda's sized types to record that subterms are smaller than the
surrounding term for the termination checker.

@@

This definition uses \AgdaFunction{$\dotto$}, which, analogously to
\AgdaFunction{$\dottimes$}, is an index-preserving version of the function
space.
We take \AgdaFunction{$\dotto$} to handle $n$ many indices --- in this
case two (the context and the type).
The notation
\AgdaFunction{$\forall[$}\AgdaSpace{}\AgdaBound{T}\AgdaSpace{}\AgdaFunction{]}
stands for
\AgdaSymbol{$\forall$}\AgdaSpace{}\AgdaSymbol{\{}%
\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$}%
\AgdaSymbol{\}}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaBound{T}%
\AgdaSpace{}%
\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$},
where \AgdaBound{T} is a type family with $n$ many indices.

@@

Finally, I give an example term \AgdaFunction{myGraph}.
As I said at the end of \cref{sec:terms}, writing a term out in full is tedious,
so I instead choose to use the machinery I discuss in detail in
\cref{sec:usage-elaborator}.
The \AgdaFunction{elab-unique} tool elaborates a well typed term into a well
typed and usage-correct term as long as it can infer assignments of usage
annotations that satisfy the constraints.
I add the prefix \texttt{u} to signify the unannotated (just well typed) terms.
All of the checks required by the elaboration procedure are done at Agda's type
checking time, so we know that the program listed below can be elaborated by
virtue of the whole Agda definition being type-checked.

@@

\subsection{The system $\mu\tilde\mu$}
I encode a usage-annotated version of System $L$/the
$\mu\tilde\mu$-calculus~\citep{CH00} --- a syntax for classical logic --- in
such a way that contexts capture the undistinguished parts of the sequent.
As such, the generic substitution lemma we get in \cref{sec:kit-to-sem} is the
form of substitution required in standard $\mu\tilde\mu$-calculus metatheory.
Though the $\mu\tilde\mu$-calculus is originally described as a sequent
calculus~\citep{CH00}, I use the techniques of
\citet[p.~12]{herbelin-hab} and \citet{LC06} to present it using hypothetical
judgements, thus giving a notion of \emph{variable} to the system.

>>>>

\section{Terms of a \AgdaRecord{System}}\label{sec:terms-of-system}

@@

To specify the meaning of descriptions, we assume some
\AgdaBound{X}\AgdaSpace{}\AgdaSymbol:\AgdaSpace{}\AgdaFunction{ExtOpenFam}%
\AgdaSpace{}\AgdaSymbol{\_},
% \ExecuteMetaData[\Interpretationtex]{X},
over which we form one layer of syntax, using the function
\AgdaFunction{$\llbracket$\_$\rrbracket$p} that interprets
\AgdaDatatype{Premises} defined below.  The first argument to
\AgdaBound{X} is the new variables bound by this layer of syntax, as
exemplified in the first clause of
\AgdaFunction{$\llbracket$\_$\rrbracket$p}.  The second argument is
the context containing the variables being carried over from the
previous layer.  Notice that this is not, in general, the same as the
context from the previous layer, because the usage annotations may
have been changed by connectives like
\AgdaInductiveConstructor{\_`$*$\_} and
\AgdaInductiveConstructor{\_`$\cdot$\_}.  The third argument is the
type of subterm required.

The remainder of the clauses of \AgdaFunction{$\llbracket$\_$\rrbracket$p}
are given by the bunched connectives, as listed below.

@@

The most obvious way to make an
\AgdaBound{X}\AgdaSpace{}\AgdaSymbol:\AgdaSpace{}\AgdaFunction{ExtOpenFam}%
\AgdaSpace{}\AgdaSymbol{\_}
is to use some existing
\AgdaFunction{OpenFam} on an extended context.
I define \AgdaFunction{Scope} to do this: take the new variables
\AgdaBound{$\Delta$}, concatenate them onto the existing context
\AgdaBound{$\Gamma$}, and pass the extended context onto the judgement
\AgdaBound{T}.

@@

I use \AgdaFunction{Scope} to deal with new variables in syntax.
Terms resemble the free monad over a layer-of-syntax functor, though
that picture is complicated by variable binding.  A term is either a
variable or a use of a logical rule together with terms for each of
the required subterms.
The \AgdaFunction{Size} argument \AgdaBound{sz} is a use of
Agda's sized types to record that subterms are smaller than the
surrounding term for the termination checker.

@@

%This definition uses \AgdaFunction{$\dotto$}, which, analogously to
%\AgdaFunction{$\dottimes$}, is an index-preserving version of the function
%space.
%We take \AgdaFunction{$\dotto$} to handle $n$ many indices --- in this
%case two (the context and the type).
%The notation
%\AgdaFunction{$\forall[$}\AgdaSpace{}\AgdaBound{T}\AgdaSpace{}\AgdaFunction{]}
%stands for
%\AgdaSymbol{$\forall$}\AgdaSpace{}\AgdaSymbol{\{}%
%\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$}%
%\AgdaSymbol{\}}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaBound{T}%
%\AgdaSpace{}%
%\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$},
%where \AgdaBound{T} is a type family with $n$ many indices.

@@

Finally, I give an example term \AgdaFunction{myGraph}.
As I said at the end of \cref{sec:terms-of-system}, writing a term out in full
is tedious, so I instead choose to use the machinery I discuss in detail in
\cref{sec:usage-elaborator}.
The \AgdaFunction{elab-unique} tool elaborates a well typed term into a well
typed and usage-correct term as long as it can infer assignments of usage
annotations that satisfy the constraints.
I add the prefix \texttt{u} to signify the unannotated (just well typed) terms.
All of the checks required by the elaboration procedure are done at Agda's type
checking time, so we know that the program listed below can be elaborated by
virtue of the whole Agda definition being type-checked.

@@

\subsection{The system $\mu\tilde\mu$}\label{sec:mu-mu-tilde}
I encode a usage-annotated version of System $L$/the
$\mu\tilde\mu$-calculus~\citep{CH00} --- a syntax for classical logic --- in
such a way that contexts capture the undistinguished parts of the sequent.
As such, the generic substitution lemma we get in \cref{sec:kit-to-sem} is the
form of substitution required in standard $\mu\tilde\mu$-calculus metatheory.
Though the $\mu\tilde\mu$-calculus is originally described as a sequent
calculus~\citep{CH00}, I use the techniques of
\citet[p.~12]{herbelin-hab} and \citet{LC06} to present it using hypothetical
judgements, thus giving a notion of \emph{variable} to the system.
